High-level cognition is supported by information-rich but compressible brain activity patterns

### 2024-03-12 paper/main.tex
added autocorrelation analysis
add spatial_autocorrelations.pkl to dropbox:   - load locally if possible   - else download from dropbox   - if failure, compute and store locally
the pkl file is 1.2 GB, so it's too large to store directly on GitHub

\newcommand{\varExplained}{S1}
\newcommand{\topTerms}{S2}
\newcommand{\componentBrains}{S3}
\newcommand{\topicContrasts}{S4}
\newcommand{\neurosynthFull}{S5}

@@

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns and networks~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10,
SizeEtal18, SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11,
AdacEtal12, BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04,
DhamEtal08, KorzEtal08, BrovEtal04, LynnBass21}. Correlations or causal
associations between different brain regions simultaneously imply that
full-brain activity patterns will be compressible and also that those activity
patterns will contain redundancies. For example, the extent to which activity
patterns at one brain area can be inferred or predicted from activity patterns
at other areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then a
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region. Whereas compressible brain
patterns are robust to spatial signal corruption, high versus low
informativeness reflects a similar (though dissociable; e.g.,
Fig.~\ref{fig:information-compression}) tradeoff between expressiveness and
robustness of \textit{temporal} patterns. Highly informative brain patterns (by
our measure; i.e., patterns that yield greater temporal decoding accuracy) are
expressive about ongoing experiences or cognitive states, since each moment's
pattern is reliably distinguishable from other moments' patterns. However, when
each moment's pattern is unique, brain activity becomes less robust to temporal
signal corruption. Our finding that brain activity patterns becomes more
informative (i.e., less robust to temporal signal corruption) and compressible
(i.e., more robust to spatial signal corruption) when cognitive engagement is
higher suggests that our brain may optimize its activity patterns to prioritize
either temporal or spatial robustness, according to task demands.

@@

One potential limitation of our work concerns how our measure of
informativeness might generalize across different tasks, cognitive
representations, and processes. Our use of across-participant temporal decoding
accuracy as a proxy for informativeness is motivated in part by prior work that
introduced across-participant similarity (in time-varying response to a
stimulus) as a means of identifying stimulus-driven brain activity
patterns~\citep{SimoEtal16}. Intuitively, only activity patterns that are
driven by the stimulus would be expected to synchronize (i.e., be time-locked
to the stimulus) across participants. This approach implicitly removes
idiosyncratic responses (e.g., neural patterns that are \textit{not} similar
across people). However, there are also some published examples, including in
our own prior work, that indicate that some types of stimulus-evoked activity
will be missed by across-participant comparisons. For example we have reported
how brain regions like the ventromedial prefrontal cortex (vmPFC) show
stimulus-driven responses that are, for the most part, \textit{not} similar
across people. In that paper (and drawing on other work), we suggest that the
vmPFC seems to represent or support highly idiosyncratic internal states, like
affective responses. Although we would consider the vmPFC to be a
``high-level'' region (e.g., we consider affect to be a relatively high-level
aspect of cognition), the measure of informativeness that we used in our
current study would identify regions like the vmPFC as having \textit{low}
informativeness. This is because across-participant decoding accuracy (our
proxy measure for informativeness) will only be high for representations or
responses that are common across people.

>>>>

\newcommand{\autocorr}{S1}
\newcommand{\varExplained}{S2}
\newcommand{\topTerms}{S3}
\newcommand{\componentBrains}{S4}
\newcommand{\topicContrasts}{S5}
\newcommand{\neurosynthFull}{S6}

@@

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns and networks~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10,
SizeEtal18, SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11,
AdacEtal12, BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04,
DhamEtal08, KorzEtal08, BrovEtal04, LynnBass21, ShinnEtal23}. Correlations or causal
associations between different brain regions simultaneously imply that
full-brain activity patterns will be compressible and also that those activity
patterns will contain redundancies. For example, the extent to which activity
patterns at one brain area can be inferred or predicted from activity patterns
at other areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then a
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region. Whereas compressible brain
patterns are robust to spatial signal corruption, high versus low
informativeness reflects a similar (though dissociable; e.g.,
Fig.~\ref{fig:information-compression}) tradeoff between expressiveness and
robustness of \textit{temporal} patterns. Highly informative brain patterns (by
our measure; i.e., patterns that yield greater temporal decoding accuracy) are
expressive about ongoing experiences or cognitive states, since each moment's
pattern is reliably distinguishable from other moments' patterns. However, when
each moment's pattern is unique, brain activity becomes less robust to temporal
signal corruption. Our finding that brain activity patterns becomes more
informative (i.e., less robust to temporal signal corruption) and compressible
(i.e., more robust to spatial signal corruption) when cognitive engagement is
higher suggests that our brain may optimize its activity patterns to prioritize
either temporal or spatial robustness, according to task demands.

@@

One potential limitation of our work concerns how our measure of
informativeness might generalize across different tasks, cognitive
representations, and processes. As in many neuroimaing studies, our findings
may also be influenced by our preprocessing decisions, which could in turn
affect temporal and spatial correlations in the data. Our use of
across-participant temporal decoding accuracy as a proxy for informativeness is
motivated in part by prior work that introduced across-participant similarity
(in time-varying response to a stimulus) as a means of identifying
stimulus-driven brain activity patterns~\citep{SimoEtal16}. Intuitively, only
activity patterns that are driven by the stimulus would be expected to
synchronize (i.e., be time-locked to the stimulus) across participants. This
approach implicitly removes idiosyncratic responses (e.g., neural patterns that
are \textit{not} similar across people). However, there are also some published
examples, including in our own prior work, that indicate that some types of
stimulus-evoked activity will be missed by across-participant comparisons. For
example we have reported how brain regions like the ventromedial prefrontal
cortex (vmPFC) show stimulus-driven responses that are, for the most part,
\textit{not} similar across people~\citep{ChanEtal21a}. In that paper (and
drawing on other work), we suggest that the vmPFC seems to represent or support
highly idiosyncratic internal states, like affective responses. Although we
would consider the vmPFC to be a ``high-level'' region (e.g., we consider
affect to be a relatively high-level aspect of cognition), the measure of
informativeness that we used in our current study would identify regions like
the vmPFC as having \textit{low} informativeness. This is because
across-participant decoding accuracy (our proxy measure for informativeness)
will only be high for representations or responses that are common across
people.
